---
title: "What Makes Oregon Wine Unique?"
author: "Jameson Watts, Ph.D."
date: "04/16/2020"
output:
  ioslides_presentation:
    widescreen: yes
    incremental: yes
    df_print: kable
    fig_width: 10
---

<style>
strong{
  color: #018080;
}
table.rmdtable th {
    color: black;
    background: #dbc4cA;
}
table.rmdtable td {
    padding: 0.5em 0.5em;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)

library(tidyverse)
library(tidytext)
library(scales)

wine <- read_csv("winemag-data.csv") %>% 
  filter(!is.na(price)) %>%
  mutate(year = as.numeric(str_extract(title,"(\\d{4})")))

data(stop_words)
wtxt <- wine %>% 
  filter(variety=="Pinot Noir") %>% 
  unnest_tokens(word, description) %>% 
  anti_join(stop_words) %>% 
  filter(str_detect(string = word, pattern = "[a-z+]")) %>%  # get rid weird non alphas
  filter(str_length(word)>3) %>%  # get rid of strings shorter than 3 characters
  group_by(variety, word) %>% 
  mutate(total=n()) %>% 
  ungroup()

west_coast <- c("Washington","Oregon","California")
pinots <- c("Burgundy","California","Oregon")

library(wesanderson)
jameson <- function(){
  list(
    theme_minimal() +
    theme(
      text = element_text(family = "Palatino", color = "gray25"),
      plot.subtitle = element_text(size = 12),
      plot.caption = element_text(color = "gray30"),
      plot.background = element_rect(fill = "gray95"),
      plot.margin = unit(c(5, 10, 5, 10), units = "mm")
    ),
    scale_color_manual(values = wes_palette(n=10, name="Moonrise3", type = "continuous")),
    scale_fill_manual(values = wes_palette(n=10, name="Moonrise3", type = "continuous"))
  )
}
```


## Pinot Noir !!!

### Right??
\
\
But wait ...is there evidence?

## Top 5 wine varieties on the west coast
```{r}
top_5 <- (wine %>%
  filter(province %in% west_coast) %>%
  count(variety) %>% 
  top_n(5,n))$variety 

wine %>% 
  filter(province %in% west_coast) %>%
  filter(variety %in% top_5) %>% 
  ggplot(aes(x=province, fill=variety)) +
  geom_bar()+
  jameson()
```

## Wines at 98 points or higher on the west coast
```{r}
wine %>%
  filter(points >= 98) %>%
  filter(province %in% west_coast) %>%
  ggplot(aes(x=province, fill=variety)) +
  geom_bar()+
  jameson()
```

## Oregon's Top wines 
```{r}
wine %>% 
  filter(province == "Oregon") %>% 
  filter(points >= 97) %>% 
  arrange(desc(points),desc(price)) %>% 
  select(winery,variety,points,price)
```

# So what really makes us unique?



## Different from California Pinot...

```{r}
wtxt %>% 
    filter(province=="Oregon" | province=="California") %>% 
    filter(variety=="Pinot Noir") %>% 
    filter(!(word %in% c("wine","pinot","drink","noir","vineyard","palate","notes","flavors","bottling"))) %>% 
    filter(total > 400) %>% 
    group_by(province, word) %>%
    count() %>% 
    group_by(province) %>% 
    mutate(proportion = n / sum(n)) %>% 
    pivot_wider(id_cols = word, names_from = province, values_from = proportion) %>% 
    ggplot(aes(x = Oregon, y = California, color = abs(Oregon - California))) +
    geom_abline(color = "gray40", lty = 2) +
    geom_jitter(alpha = 0.1, size = 2.5, width = 0.3, height = 0.3) +
    geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
    scale_x_log10(labels = percent_format()) +
    scale_y_log10(labels = percent_format()) +
    scale_color_gradient(limits = c(0, 0.001), low = "darkslategray4", high = "gray75") +
    theme(legend.position="none") +
    labs(x = "Oregon", y = "California", title = "Words describing Pinot Noir from California and Oregon")
```

## Different from French Burgundy
```{r, warning=FALSE}
wtxt %>% 
  filter(province=="Oregon" | province=="Burgundy") %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(!(word %in% c("wine","pinot","drink","noir","vineyard","palate","notes","flavors","bottling","bottle","finish"))) %>% 
  filter(total > 400) %>% 
  group_by(province, word) %>%
  count() %>% 
  group_by(province) %>% 
  mutate(proportion = n / sum(n)) %>% 
  pivot_wider(id_cols = word, names_from = province, values_from = proportion) %>% 
  ggplot(aes(x = Oregon, y = Burgundy, color = abs(Oregon - Burgundy))) +
  geom_abline(color = "gray40", lty = 2) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.3, height = 0.3) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  scale_color_gradient(limits = c(0, 0.001), low = "darkslategray4", high = "gray75") +
  theme(legend.position="none") +
  labs(x = "Oregon", y = "Burgundy", title = "Describing Pinot Noir from Burgundy and Oregon")
```

## Unique Oregon pinot descriptors (compared to California)

```{r}
dtxt <- wtxt %>% 
  filter(province=="Oregon" | province=="California") %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(!(word %in% c("wine","pinot","drink","noir","vineyard","palate","notes","flavors","bottling","bottle","finish"))) %>% 
  filter(total > 400) %>% 
  group_by(province, word) %>%
  count() %>% 
  group_by(province) %>% 
  mutate(proportion = n / sum(n)) %>% 
  pivot_wider(id_cols = word, names_from = province, values_from = proportion) %>% 
  mutate(diff=Oregon-California) 

dtxt %>%
  top_n(25, diff) %>%
  mutate(word = reorder(word, diff)) %>%
  ggplot(aes(word, diff)) +
  geom_col() +
  xlab(NULL) +
  coord_flip()+
  jameson()
```

## Unique Oregon pinot descriptors (compared to Burgundy)

```{r}
dtxt <- wtxt %>% 
  filter(province=="Oregon" | province=="Burgundy") %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(!(word %in% c("wine","pinot","drink","noir","vineyard","palate","notes","flavors","bottling","bottle","finish"))) %>% 
  filter(total > 400) %>% 
  group_by(province, word) %>%
  count() %>% 
  group_by(province) %>% 
  mutate(proportion = n / sum(n)) %>% 
  pivot_wider(id_cols = word, names_from = province, values_from = proportion) %>% 
  mutate(diff=Oregon-Burgundy) 

dtxt %>%
  top_n(25, diff) %>%
  mutate(word = reorder(word, diff)) %>%
  ggplot(aes(word, diff)) +
  geom_col() +
  xlab(NULL) +
  coord_flip()+
  jameson()
```

## Rating oregon flavors

```{r}
wine %>%
  filter(variety=="Pinot Noir") %>% 
  mutate(character=
    case_when(
      str_detect(description,"fruit") ~ 'fruit',
      str_detect(description,"chocolate") ~ 'chocolate',
      str_detect(description,"tart") ~ 'tart'
    )
  ) %>% 
  filter(!is.na(character)) %>% 
  ggplot(aes(points, fill=character))+
    geom_density(alpha=.5)+
  jameson()
```

## A different view...

```{r}
wine %>%
  filter(variety=="Pinot Noir") %>% 
  mutate(character=
    case_when(
      str_detect(description,"fruit") ~ 'fruit',
      str_detect(description,"chocolate") ~ 'chocolate',
      str_detect(description,"tart") ~ 'tart'
    )
  ) %>% 
  filter(!is.na(character)) %>% 
  ggplot(aes(character, points, fill=character))+
    geom_boxplot()+
    jameson()
```

## Words with highest prices from Oregon Pinots
```{r}
wtxt %>% 
  filter(province=="Oregon") %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(!(word %in% c("wine","pinot","drink","noir","vineyard","palate","notes","flavors","bottling","bottle","finish"))) %>% 
  filter(total > 400) %>% 
  group_by(word) %>%
  summarize(avg_points = mean(points), avg_price = mean(price)) %>% 
  arrange(desc(avg_price), desc(avg_points)) %>% 
  slice(1:10)
  
```

## Maybe some of these terms hang together?

Let's run a principle component analysis.

```{r}
pine <- wtxt %>% 
  filter(province=="Oregon") %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(!(word %in% c("wine","pinot","drink","noir","vineyard","palate","notes","flavors","bottling","bottle","finish"))) %>% 
  filter(total > 400) %>% 
  count(id, word) %>% 
  group_by(id) %>% 
  mutate(exists = if_else(n>0,1,0)) %>% 
  ungroup %>% 
  pivot_wider(id_cols = id, names_from = word, values_from = exists, values_fill = c(exists=0)) %>% 
  select(-id)

pr_wine <- prcomp(x = pine, scale = T, center = T,rank. = 4)
screeplot(pr_wine, type = "lines")
```

## Principle Component 1

```{r}
rownames_to_column(as.data.frame(pr_wine$rotation),var = "word") %>% 
  mutate(abs_load = abs(PC1)) %>% 
  arrange(desc(abs_load)) %>% 
  select(word, PC1) %>% 
  slice(1:5)
```

## Principle Component 2

```{r}
rownames_to_column(as.data.frame(pr_wine$rotation),var = "word") %>% 
  mutate(abs_load = abs(PC2)) %>% 
  arrange(desc(abs_load)) %>% 
  select(word, PC2) %>% 
  slice(1:5)
```

## Principle Component 3

```{r}
rownames_to_column(as.data.frame(pr_wine$rotation),var = "word") %>% 
  mutate(abs_load = abs(PC3)) %>% 
  arrange(desc(abs_load)) %>% 
  select(word, PC3) %>% 
  slice(1:5)
```

## Principle Component 4

```{r}
rownames_to_column(as.data.frame(pr_wine$rotation),var = "word") %>% 
  mutate(abs_load = abs(PC4)) %>% 
  arrange(desc(abs_load)) %>% 
  select(word, PC4) %>% 
  slice(1:5)  
```

# Beyond words...

## Variation in points by province

```{r}
wine %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(province %in% pinots) %>% 
  ggplot(aes(x=province, y=points, fill=province))+
  geom_boxplot()+
  jameson()
  
```

## More consistent Pinots

```{r}
wine %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(province %in% pinots) %>% 
  ggplot(aes(x=points, fill=province))+
  geom_density(alpha=0.5)+
  jameson()
  
```

## Generally more for your money

```{r}
wine %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(province %in% pinots) %>% 
  mutate(points_per_dollar= points/price) %>% 
  ggplot(aes(x=province, y=log(points_per_dollar), fill=province))+
  geom_boxplot()+
  jameson()
  
```

## Average points per year for Pinot Noir

```{r}
wine %>% 
  filter(province %in% pinots) %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(year >= 2000 & year <= 2015) %>% 
  group_by(year, province) %>% 
  summarize(avg_points = mean(points)) %>% 
  ggplot(aes(x=as.integer(year), y=avg_points, color=province)) +
    geom_line()+
  jameson()
```

## Points per $$ for Pinot Noir

```{r}
wine %>% 
  filter(variety=="Pinot Noir") %>% 
  filter(province %in% pinots) %>% 
  ggplot(aes(x=points, y=log(price), color=province))+
  geom_point(alpha=0.1,show.legend = FALSE)+
  geom_smooth(method = lm,show.legend = FALSE) +
  facet_wrap(~ province, scales="fixed")+
  jameson()
```

## More pricing power than California

Increases in points affect wines from the different regions differently.

```{r}
pe = list()
for(p in pinots){
  model <- lm(log(price)~points, filter(wine,province==p, variety=="Pinot Noir"))
  pct <- round((exp(coef(model)["points"]) - 1) * 100,2)
  pe[[p]] <- str_c("For **",p,"**, a 1 point ratings increase leads to a **",pct,"%** increase in price.")
}
```

- `r pe$Burgundy`
- `r pe$California`
- `r pe$Oregon`

## Pricing power by flavor

```{r}
pe = list()
flavors = c("fruit","chocolate","tart")
for(f in flavors){
  model <- lm(log(price)~points, filter(wine,province=="Oregon", variety=="Pinot Noir", str_detect(description,f)))
  pct <- round((exp(coef(model)["points"]) - 1) * 100,2)
  pe[[f]] <- str_c("For Oregon Pinots with notes of **",f,"**, a 1 point ratings increase leads to a **",pct,"%** increase in price.")
}
```

- `r pe$fruit`
- `r pe$chocolate`
- `r pe$tart`

## Pricing power by type of berry
```{r}
pe = list()
flavors = c("cherry", "blackberry", "strawberry")
for(f in flavors){
  model <- lm(log(price)~points, filter(wine,province=="Oregon", variety=="Pinot Noir", str_detect(description,f)))
  pct <- round((exp(coef(model)["points"]) - 1) * 100,2)
  pe[[f]] <- str_c("For Oregon Pinots with notes of **",f,"**, a 1 point ratings increase leads to a **",pct,"%** increase in price.")
}
```

- `r pe$cherry`
- `r pe$blackberry`
- `r pe$strawberry`

## Summary

Oregon Pinot Noir...

1. Is fruit forward, with hints of chocolate and berries
2. Has lower overall variance in quality
3. Has been rising in quality over time
4. Has more pricing power than California
5. Is the best bang for your buck

Recommendations:

1. Work to retain and grow status of Willamette Valley globally
2. Smaller wineries can overcome status/recognition with targeted flavor profiles

## Thank you!!

...and we're definitely better than Texas.

```{r}

wine %>% 
  filter(province %in% c("Oregon","Texas")) %>% 
  ggplot(aes(x=points, fill=province))+
  geom_density(alpha=0.3)
  
```

## Quality Oregon Pinots under $50
```{r}
wine %>% 
  filter(province == "Oregon") %>% 
  filter(variety == "Pinot Noir") %>% 
  filter(price < 50 & points >=95) %>% 
  mutate(BFYB = round(points/price,1)) %>% 
  select(title, points, price, BFYB) %>% 
  arrange(desc(points), price)
```

## References

- https://www.kaggle.com/zynicide/wine-reviews
- https://www.tidyverse.org
- https://www.tidytextmining.com/
